An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria

### 2022-01-14 2022/OM&S/paper.tex
before submission

\begin{keywords}
Multi-objective optimization; global optimization; dimensionality reduction; search information; nature-inspired metaheuristics; numerical comparison
\end{keywords}

@@

The simplest way to build an approximation of the Pareto set consist in immersion in the feasible domain of problem's parameters a uniform grid (deterministic or random) and selection of all non-dominated nodes of the grid. Unfortunately, this universal approach is too costly because it requires in the worst case pairwise comparison of all the grid nodes and does not take into account the properties of the problem for reducing the computational costs.

@@

As for other methods of solving MCO problems it is necessary to pay attention to the important class of metaheuristic nature-inspired algorithms. Many of them are nature-inspired, i.e.,  are  based on modeling physical processes (for example, simulated annealing \cite{Locatelli2002,Aarts2014}) or mainly behavior of biological agents including evolutionary \cite{Price2005,Coello2007} and as their part genetic algorithms \cite{Ruiz2015,Deb2002,Zitzler2001} and particle swarm methods modeling the swarming intelligence of bees, fireflies, birds, ants, etc. \cite{Mostaghim2007,Nebro2009,Durillo2010}.

Consideration of different approaches to solving MCO problems and a relevant references can be found, for example, in monographs and papers \cite{Miettinen1999,Ehrgott2005,Zhou2011,Nedjah2015,Pardalos2017}.
In this paper we consider multiextremal black-box MCO problems and propose a new scheme of solving these problems. This scheme is based on maximum convolution, dimensionality reduction by means of Peano curves, information-statistical global optimization algorithm used for the first time in MCO problems. Moreover, in the framework of the scheme the information already obtained in the course of solving scalar subproblems of convolution optimization can be utilized for solving a new convolution subproblem. Briefly, the way is as follows. Before beginning the optimization of a new convolution we have at our disposal values of criteria computed already and now can recalculate them to values of the current convolution taking into account these values as initial information for optimization algorithm. This procedure accelerates significantly the optimization process.

@@

In the next section propose a modified information-statistical algorithm with accelerated convergence and its application to searching for Pareto optimal solutions in multiextremal MCO problems.

@@

In this section we consider the MCO problem under additional assumption according to which the criteria $w_i(x), \; 1 \leq i \leq k,$ are black-box functions and satisfy in the domain $Q$ the Lipschitz condition
\begin{equation}\label{lip}
\left| w_i(x) - w_i(y)\right| \leq L_i \left\| x-y \right\|, \; x,y, \in Q, \; 1 \leq i \leq k,
\end{equation}
with corresponding Lipchitz constants $L_i > 0$ that are, as a rule, unknown. Here $\left\|*\right\|$   denotes the Euclidean norm in $R^N$.

@@

For solving the problems of the family a global optimization technique based on reducing the multidimensional problem (\ref{conv_problem}) to an equivalent univariate subproblem \cite{Strongin2000,Pinter1996,Zhigljavsky2008,Sergeyev2013,PaulaviciusZilinskas2014,Sergeyev2017} in combination with one-dimensional information algorithm with accelerated convergence.

@@

To simplify the description of the technique let us present the problems (\ref{conv_problem}) in the following unified form
\begin{equation}\label{problem_f}
f(x) \rightarrow \min, \; x \in Q,
\end{equation}
where 
\[
f(x) = \max_{1 \leq i \leq k}{\lambda_i w_i(x)} + \gamma \sum_{i=1}^k{\lambda_i w_i(x)}
\]
and $f(x)$ meets the Lipschitz condition 
\begin{equation}\label{lip_1}
\left| f(x) - f(y)\right| \leq L \left\| x-y \right\|, \; x,y, \in Q,
\end{equation}
with the constant $L>0$.

@@

\begin{enumerate}
\item 
The points $t^1, \dots, t^n$ of completed trials and boundary points of the search interval $[\alpha, \beta]$ are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}\label{eq:15}
t_1 = \alpha < t_1 < \dots < t_{n-1} < t_n = \beta.
\end{equation}
The values $v_s = \varphi(t_s)$ are juxtaposed to the points  $t_s$, $1 \leq s \leq n$.
\item 
A real values $R(q)$ are assigned to all the intervals $(t_{q-1}, t_q)$, $2 \leq q \leq n$, formed by neighboring points from (\ref{eq:15}). $R(q)$ is called \textit{the characteristic} of the interval $(t_{q-1}, t_q)$ and is computed in accordance with the following rule.

If number of completed trials is not exactly divisible by $T$, where \textit{global-to-local ratio} $T$ is a predefined number, then
\begin{equation}\label{eq:16}
R(q) = B(q),
\end{equation}
otherwise
\begin{equation}\label{eq:17}
R(q) = B(q)\left( \frac {\sqrt{(v_q - \varphi^*(n))(v_{q-1} - \varphi^*(n))}}{M} +(1.5)^{-\sigma}\right)^{-1}.
\end{equation}
Here
\begin{gather}\label{eq:18}
B(q) = (t_s - t_{s-1})^{1/N} + \frac{(v_q - v_{q-1})^2} {m^2 (t_s - t_{s-1})^{1/N}} -2\frac{v_{q-1} + v_q}{m},\\
\varphi^*(n) = \min_{1\leq j \leq n} {\varphi(t^j)}
\end{gather}
is minimal computed value of the objective function $\varphi(t)$ at trial points,
\begin{equation}\label{eq:19}
M = \max_{2 \leq  s \leq n} { \frac{|v_s - v_{s-1}|}{(t_s - t_{s-1})^{1/N}} }
\end{equation}
is underestimation of the H{\"o}lder constant obtained according to the trial results,
\begin{equation}\label{eq:20}
m = 
\begin{cases}
\begin{matrix}
rM, & M>0, \\
1, & M= 0,
\end{matrix}
\end{cases}
\end{equation}
is improved estimation of the H{\"o}lder constant due to method's parameter $r>1$, $\sigma$ is some integer parameter that influences the speed of convergence to the global minimum.

@@

Like its prototype, ARL can apply as the stopping condition the inequality
\begin{equation}\label{eq:23}
(t_p-t_{p-1})^{1/N} < \varepsilon,
\end{equation}
i.e., completes the search when the length of the interval with maximal characteristic (\ref{eq:21}) becomes less than the predefined accuracy $\varepsilon > 0$.

@@

In the course of optimization the algorithm builds the trial points $t^1,...,t^n$ which correspond to the points $x^1 = x(t^1),...,x^n=x(t^n)$ in the domain $Q$ . At these points the values $v^j = \Phi_\lambda(x^j), \; 1 \leq j \leq n,$ of the function $\Phi_\lambda(x)$ from (\ref{conv_problem}) are computed as values $\varphi(t^j)$ of the function $\varphi_t$  minimized by ALR.

@@

This information can be used for decreasing the number of trials during solving the problem (\ref{conv_problem}) with other weight vector $\lambda$. Indeed, if we take the weight vector $\mu = (\mu_1,...,\mu_k) \neq \lambda$, it is easy on the base of matrix $A$ to calculate values $\Phi_\mu(x^j), \; 1 \leq j \leq n,$ at points $x^1 = x(t^1),...,x^n=x(t^n)$, i.e. values $\varphi(t^1),...,\varphi(t^n)$ of the function $\varphi(t)$ in the problem (\ref{problem_fi}) being the reduced problem (\ref{conv_problem}) with weight vector $\mu$. But in this case in ALR we can take points $t^1,...,t^n$ as points of initial trials in which values $ \varphi(t^j) = \Phi_\mu(x^j), \; 1 \leq j \leq n,$ have been already computed and continue optimization following the rules 1--4 of the algorithm. In the course of continuation, ALR will generate new points $t^q$ in which the values $\Phi_\mu(x(t^j))$, and, therefore, values $w_s(x(t^j)), \; 1 \leq s \leq k,$ will be evaluated. The latters can be added to the matrix $A$ and be taken into account in the same manner for solving other problems (\ref{conv_problem}).

@@

In total, three groups of experiments have been performed. Each group contained a series of MCO problems of the same dimension and number of criteria. Since all GKLS functions have a global minimum value equal to $-1$, the partial  criteria $w_i(x), \; 1 \leq i \leq k,$  increased by a positive constant to ensure the requirements (\ref{positive}).

>>>>

\begin{keywords}
Multiobjective optimization; global optimization; dimensionality reduction; search information; nature-inspired metaheuristics; numerical comparison
\end{keywords}

@@

The simplest way to build an approximation of the Pareto set consists in immersion in the feasible domain of problem's parameters a uniform grid (deterministic or random) and in selection of all non-dominated nodes of the grid. Unfortunately, this universal approach is too costly because it requires in the worst case pairwise comparison of all the grid nodes and does not take into account the properties of the problem for reducing the computational costs.

@@

Note that special classes of multiextremal optimization problems arise when the criteria can be described analytically. In this case one can apply efficient methods based on d.c. optimization approach \cite{Strekalovsky2001,Strekalovsky2003,Strekalovsky2020}. However, this paper considers a completely different case, when the analytical expressions for the functions are not known. This situation is typical for finding the best decisions in applied problems in which the evaluation of criteria requires numerical simulation.

As for other methods of solving MCO problems it is necessary to pay attention to the important class of metaheuristic algorithms. Many of them are nature-inspired, i.e.,  are  based on modeling physical processes (for example, simulated annealing \cite{Locatelli2002,Aarts2014}) or mainly behavior of biological agents including evolutionary \cite{Price2005,Coello2007} and as their part genetic algorithms \cite{Ruiz2015,Deb2002,Zitzler2001} and particle swarm methods modeling the swarming intelligence of bees, fireflies, birds, ants, etc. \cite{Mostaghim2007,Nebro2009,Durillo2010}.

Consideration of different approaches to solving MCO problems and a relevant references can be found, for example, in monographs and papers \cite{Miettinen1999,Ehrgott2005,Zhou2011,Nedjah2015,Pardalos2017}.
In this paper we consider multiextremal black-box MCO problems and propose a new scheme of solving these problems. This scheme is based on maximum convolution, dimensionality reduction by means of Peano curves, modified information-statistical global optimization algorithm used for the first time in MCO problems. Moreover, in the framework of the scheme the information already obtained in the course of solving scalar subproblems of convolution optimization can be utilized for solving a new convolution subproblem. Briefly, the way is as follows. Before beginning the optimization of a new convolution we have at our disposal values of criteria computed earlier and now can recalculate them to values of the current convolution taking into account these values as initial information for optimization algorithm. This procedure accelerates significantly the optimization process.

@@

In the next section we propose a modified information-statistical algorithm with accelerated convergence and its application to searching for Pareto optimal solutions in multiextremal MCO problems.

@@

In this section we consider the MCO problem under additional assumptions according to which the criteria $w_i(x), \; 1 \leq i \leq k,$ are black-box functions and satisfy in the domain $Q$ the Lipschitz condition
\begin{equation}\label{lip}
\left| w_i(x) - w_i(y)\right| \leq L_i \left\| x-y \right\|, \; x,y, \in Q, \; 1 \leq i \leq k,
\end{equation}
with corresponding Lipchitz constants $L_i > 0$ that are, as a rule, unknown. Here $\left\|*\right\|$   denotes the Euclidean norm in $R^N$.

@@

For solving the problems of the family, a global optimization technique based on reducing the multidimensional problem (\ref{conv_problem}) to an equivalent univariate subproblem \cite{Strongin2000,Pinter1996,Zhigljavsky2008,Sergeyev2013,PaulaviciusZilinskas2014,Sergeyev2017} in combination with one-dimensional information algorithm with accelerated convergence is proposed.

@@

To simplify the description let us present the problems (\ref{conv_problem}) in the following unified form
\begin{equation}\label{problem_f}
f(x) \rightarrow \min, \; x \in Q,
\end{equation}
where 
\[
f(x) = \max_{1 \leq i \leq k}{\lambda_i w_i(x)} + \gamma \sum_{i=1}^k{\lambda_i w_i(x)}
\]
and $f(x)$ meets the Lipschitz condition 
\begin{equation}\label{lip_1}
\left| f(x) - f(y)\right| \leq L \left\| x-y \right\|, \; x,y, \in Q,
\end{equation}
with the constant $L>0$.

@@

\begin{enumerate}
\item 
The points $t^1, \dots, t^n$ of completed trials are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}\label{eq:15}
t_1 = \alpha < t_1 < \dots < t_{n-1} < t_n = \beta.
\end{equation}
The values $v_s = \varphi(t_s)$ are juxtaposed to the points  $t_s$, $1 \leq s \leq n$.
\item 
A real values $R(q)$ are assigned to all the intervals $(t_{q-1}, t_q)$, $2 \leq q \leq n$, formed by neighboring points from (\ref{eq:15}). $R(q)$ is called \textit{the characteristic} of the interval $(t_{q-1}, t_q)$ and is computed in accordance with the following rule.

If number of completed trials is not exactly divisible by $T$, where \textit{global-to-local ratio} $T$ is a predefined number, then
\begin{equation}\label{eq:16}
R(q) = B(q),
\end{equation}
otherwise
\begin{equation}\label{eq:17}
R(q) = B(q)\left( \frac {\sqrt{(v_q - \varphi^*(n))(v_{q-1} - \varphi^*(n))}}{M} +(1.5)^{-\sigma}\right)^{-1}.
\end{equation}
Here
\begin{gather}\label{eq:18}
B(q) = (t_s - t_{s-1})^{1/N} + \frac{(v_q - v_{q-1})^2} {m^2 (t_s - t_{s-1})^{1/N}} -2\frac{v_{q} + v_{q-1}}{m},\\
\varphi^*(n) = \min_{1\leq j \leq n} {\varphi(t^j)}
\end{gather}
is minimal computed value of the objective function $\varphi(t)$ at trial points,
\begin{equation}\label{eq:19}
M = \max_{2 \leq  s \leq n} { \frac{|v_s - v_{s-1}|}{(t_s - t_{s-1})^{1/N}} }
\end{equation}
is underestimation of the H{\"o}lder constant obtained according to the trial results,
\begin{equation}\label{eq:20}
m = 
\begin{cases}
\begin{matrix}
rM, & M>0, \\
1, & M= 0,
\end{matrix}
\end{cases}
\end{equation}
is improved estimation of the H{\"o}lder constant due to method's parameter $r>1$, $\sigma$ is some integer parameter that influences the speed of convergence to the global minimum.

@@

Like its prototype, ARL can apply as the stopping condition the inequality
\begin{equation}\label{eq:23}
(t_p-t_{p-1})^{1/N} < \varepsilon,
\end{equation}
i.e., it completes the search when the length of the interval with maximal characteristic (\ref{eq:21}) becomes less than the predefined accuracy $\varepsilon > 0$.

@@

In the course of optimization the algorithm builds the trial points $t^1,...,t^n$ which correspond to the points $x^1 = x(t^1),...,x^n=x(t^n)$ in the domain $Q$ . At these points the values $v^j = \Phi_\lambda(x^j), \; 1 \leq j \leq n,$ of the function $\Phi_\lambda(x)$ from (\ref{conv_problem}) are computed as values $\varphi(t^j)$ of the function $\varphi(t)$  minimized by ALR.

@@

This information can be used for decreasing the number of trials during solving the problem (\ref{conv_problem}) with other weight vector. Indeed, if we take the weight vector $\mu = (\mu_1,...,\mu_k) \neq \lambda$, it is easy on the base of matrix $A$ to calculate values $\Phi_\mu(x^j), \; 1 \leq j \leq n,$ at points $x^1 = x(t^1),...,x^n=x(t^n)$, i.e. values $\varphi(t^1),...,\varphi(t^n)$ of the function $\varphi(t)$ in the problem (\ref{problem_fi}) being the reduced problem (\ref{conv_problem}) with weight vector $\mu$. But in this case in ALR we can take points $t^1,...,t^n$ as points of initial trials in which values $ \varphi(t^j) = \Phi_\mu(x^j), \; 1 \leq j \leq n,$ have been already computed and continue optimization following the rules 1--4 of the algorithm. In the course of continuation, ALR will generate new points $t^q$ in which the values $\Phi_\mu(x(t^j))$, and, therefore, values $w_s(x(t^j)), \; 1 \leq s \leq k,$ will be evaluated. The latters can be added to the matrix $A$ and be taken into account in the same manner for solving other problems (\ref{conv_problem}).

@@

In total, three groups of experiments have been performed. Each group contained a series of MCO problems of the same dimension and number of criteria. Since all GKLS functions have a global minimum value equal to $-1$, the partial  criteria $w_i(x), \; 1 \leq i \leq k,$  were increased by a positive constant to ensure the requirements (\ref{positive}).
